### 2020-12-03 submission/manuscript.Rmd
First draft of Methods section

**word choice: sample vs. community**

@@

**EDIT: Two elements that are held in tension in the analysis of 16S rRNA gene sequence data is how to adequately remove PCR and sequencing artifacts and decrease the granularity of the taxonomic level that is used in the analysis.** Previous attempts have included screening for sequencing quality based on quality scores [Kozich/Edgar] followed by a polishing step based on the frequency of the sequences relative to similar sequences [Kozich/Edgar/DeBlur]. Other pipelines model the quality scores and types of errors to cluster sequences directly [Dada2]. But, as a final step many pipelines advocate for removing rare sequences from each dataset prior to outputting the sequence data as amplicon sequence variants (ASVs) [Knight/Edgar/DeBlur/Dada2]. ASVs are often clustered further to generate operational taxonomic units or phylotypes. Some pipelines remove all ASVs that appear once (i.e. singletons) [XXXX], XXXXXX [XXXXXX], XXXXX [XXXXXXX], or XXXXXX [XXXXXX] times prior to further clustering or making ecological comparisions. Notably, the mothur-based pipeline discourages the practice of removing rare sequences.

@@

**Datasets.** I collected `r nrow(table_1)` publicly available datasets that used the Illumina MiSeq platform to sequence the V4 region of the 16S rRNA gene from a variety of environments (Table 1). To insure the highest possible data quality, datasets were limited to those where the 500 cycle v2 MiSeq chemistry was used to sequence the amplicons. The paired 250 nt reads resulted in near complete 2-fold sequencing coverage of every nucleotide in the ca. 250 nt-long region. This region and sequencing platform were selected because previous work has shown that a standard data analysis pipeline in mothur results in a sequencing error rate below 0.02% [@Kozich2013]. All sequence data were obtained from the Sequence Read Archive (SRA) and processed using a standard mothur-based sequencing pipeline that resulted in ASVs as generated by the pre.cluster algorithm [@Kozich2013; @Schloss2009]. After removing poor quality and chimeric ASVs and samples that had uncharacteristically low number of sequences for the dataset, these datasets included between `r oxford_comma(n_sample_range)` samples (Figure S1). The median number of sequences for each dataset ranged between `r oxford_comma(n_med_sequence_range)` (Table 1). Strikingly, aside from the relatively small `r oxford_comma(low_fold_samples)` datasets, the difference between the sample with the fewest sequences and the sample with the most sequences for each dataset varied by between `r oxford_comma(high_fold_range, digits=1)`-fold (Table 1).

@@

**The impact of removing rare ASVs on the ability to detect statistically significant differences between treatment groups.** To test the effect of increased inter-sample variation, I randomly assigned samples to one of two treatment groups. In the first treatment group, communities were randomly sampled from the null distribution as described above. For the second treatment group, I randomly selected 10% of the ASVs in the pooled study distribution to increase their abundance by 5%. I randomly generated 100 simulated sets of treatment groups and samples. I then tested the ability to detect a difference between the two treatment groups using alpha and beta diversity metrics. The fraction of significant tests was a measurement of the statistical power to detect the difference between the treatment groups. When considering the differences in richness and diversity, the marine dataset yielded no simulated sets that were statistically significant, which was likely due to the small number of samples in the study (N=7). Among the remaining datasets, the power to detect a difference in the richness of ASVs ranged between `r oxford_comma(pull(b_base_pc_sobs,'1'), 2)` (`r oxford_comma(pull(b_base_pc_sobs,dataset))`) and between `r oxford_comma(pull(b_base_pc_shannon, '1'), 2)` (`r oxford_comma(pull(b_base_pc_shannon, dataset))`) to detect a difference in diversity when using a Wilcox test (Figure 4A). When singleton ASVs were removed, the power to detect a difference in the diversity of ASVs dropped by between `r oxford_comma(pull(b_pc_sobs, percentage_drop), 1)`% (`r oxford_comma(pull(b_pc_sobs, dataset))`) and by between `r oxford_comma(pull(b_pc_shannon, percentage_drop), 1)`% (`r oxford_comma(pull(b_pc_shannon, dataset))`; Figure 4B). The effect of removing rare ASVs on the richness of OTUs and their diversity was similar (Figure S5AB). I used the Bray-Curtis dissimilarity index to compare the simulated communities within each dataset and calculated the power to detect differences between the two simulated treatment groups using the analysis of molecular variance (also called PERMANOVA) (Figure 4C and S5C). Without removing rare sequences, the power to detect a difference between the two simulated treatment groups varied between `r oxford_comma(pull(b_base_beta, '1') %>% range(), 2)` (`r slice(b_base_beta, 1) %>% pull(dataset)` and `r oxford_comma(slice_max(b_base_beta, '1') %>% pull(dataset))`). Aside from the `r oxford_comma(unfazed_singletons)` datasets, the power to detect differences dropped by between `r oxford_comma(pull(fazed_singletons,p_drop_singletons), 1)`% (`r oxford_comma(pull(fazed_singletons, dataset))`) when singletons were removed. However, when ASVs that occurred 10 or fewer times were removed from each sample, the power to detect differences dropped by `r oxford_comma(pull(fazed_tentons, p_drop_tens), 1)`% (`r oxford_comma(pull(fazed_tentons, dataset))`); similar results were observed when ASVs were clustered into OTUs. Removing rare ASVs reduced the ability to detect simulated treatment effects using metrics commonly used to compare microbial communities.

@@

Datasets were selected because they sequenced the V4 region of the 16S rRNA gene using Illumina's MiSeq platform using the v2 chemistry with paired 250 nt reads. I identified 12 datasets that satisfied these criteria and downloaded the data directly from the SRA (Table 1 and Figure S1).

* Used the mothur pipeline to
  - Assembly of paired 250 nt reads
*

>>>>

**word choice: sample vs. community**
**word choice: dataset vs. sample**

@@

Two elements that are held in tension in the analysis of 16S rRNA gene sequence data is how to adequately remove PCR and sequencing artifacts while decreasing the granularity of the taxonomic level that is used in the analysis. Previous attempts have included screening for sequencing quality based on quality scores [Kozich/Edgar] followed by a polishing step based on the frequency of the sequences relative to similar sequences [Kozich/Edgar/DeBlur]. Other pipelines model the quality scores and types of errors to cluster sequences directly [Dada2]. But, as a final step many pipelines advocate for removing rare sequences from each dataset prior to outputting the sequence data as amplicon sequence variants (ASVs) [Knight/Edgar/DeBlur/Dada2]. ASVs are often clustered further to generate operational taxonomic units or phylotypes. Some pipelines remove all ASVs that appear once (i.e. singletons) [XXXX], XXXXXX [XXXXXX], XXXXX [XXXXXXX], or XXXXXX [XXXXXX] times prior to further clustering or making ecological comparisions. Notably, the mothur-based pipeline discourages the practice of removing rare sequences.

@@

**Datasets.** I collected `r nrow(table_1)` publicly available datasets that used the Illumina MiSeq platform to sequence the V4 region of the 16S rRNA gene from a variety of environments (Table 1). After removing poor quality and chimeric ASVs and samples that had uncharacteristically low number of sequences for the dataset, these datasets included between `r oxford_comma(n_sample_range)` samples (Figure S1). The median number of sequences for each dataset ranged between `r oxford_comma(n_med_sequence_range)` (Table 1). Strikingly, aside from the relatively small `r oxford_comma(low_fold_samples)` datasets, the difference between the sample with the fewest sequences and the sample with the most sequences for each dataset varied by between `r oxford_comma(high_fold_range, digits=1)`-fold (Table 1).

@@

**The impact of removing rare ASVs on the ability to detect statistically significant differences between treatment groups.** To test the effect of increased inter-sample variation, I randomly assigned samples to one of two treatment groups. In the first treatment group, communities were randomly sampled from the null distribution as described above. For the second treatment group, I randomly selected 10% of the ASVs in the pooled study distribution to increase their abundance by 5%. I randomly generated 100 simulated sets of treatment groups and samples. I then tested the ability to detect a difference between the two treatment groups using alpha and beta diversity metrics. The fraction of significant tests was a measurement of the statistical power to detect the difference between the treatment groups. When considering the differences in richness and diversity, the marine dataset yielded no simulated sets that were statistically significant, which was likely due to the small number of samples in the study (N=7). Among the remaining datasets, the power to detect a difference in the richness of ASVs ranged between `r oxford_comma(pull(b_base_pc_sobs,'1'), 2)` (`r oxford_comma(pull(b_base_pc_sobs,dataset))`) and between `r oxford_comma(pull(b_base_pc_shannon, '1'), 2)` (`r oxford_comma(pull(b_base_pc_shannon, dataset))`) to detect a difference in diversity when using a Wilcox test (Figure 4A). When singleton ASVs were removed, the power to detect a difference in the diversity of ASVs dropped by between `r oxford_comma(pull(b_pc_sobs, percentage_drop), 1)`% (`r oxford_comma(pull(b_pc_sobs, dataset))`) and by between `r oxford_comma(pull(b_pc_shannon, percentage_drop), 1)`% (`r oxford_comma(pull(b_pc_shannon, dataset))`; Figure 4B). The effect of removing rare ASVs on the richness of OTUs and their diversity was similar (Figure S5AB). I used the Bray-Curtis dissimilarity index to compare the simulated communities within each dataset and calculated the power to detect differences between the two simulated treatment groups using the analysis of molecular variance (Figure 4C and S5C). Without removing rare sequences, the power to detect a difference between the two simulated treatment groups varied between `r oxford_comma(pull(b_base_beta, '1') %>% range(), 2)` (`r slice(b_base_beta, 1) %>% pull(dataset)` and `r oxford_comma(slice_max(b_base_beta, '1') %>% pull(dataset))`). Aside from the `r oxford_comma(unfazed_singletons)` datasets, the power to detect differences dropped by between `r oxford_comma(pull(fazed_singletons,p_drop_singletons), 1)`% (`r oxford_comma(pull(fazed_singletons, dataset))`) when singletons were removed. However, when ASVs that occurred 10 or fewer times were removed from each sample, the power to detect differences dropped by `r oxford_comma(pull(fazed_tentons, p_drop_tens), 1)`% (`r oxford_comma(pull(fazed_tentons, dataset))`); similar results were observed when ASVs were clustered into OTUs. Removing rare ASVs reduced the ability to detect simulated treatment effects using metrics commonly used to compare microbial communities.

@@

**Data curation and analysis.** To insure the highest possible data quality, datasets were limited to those where the 500 cycle v2 MiSeq chemistry was used to sequence the amplicons. The paired 250 nt reads resulted in near complete 2-fold sequencing coverage of every nucleotide in the ca. 250 nt-long region. This region and sequencing platform were selected because previous work has shown that a standard data analysis pipeline in mothur results in a sequencing error rate below 0.02% [@Kozich2013]. All sequence data were obtained from the Sequence Read Archive (SRA) and processed using a standard mothur-based sequencing pipeline that resulted in ASVs as generated by the pre.cluster algorithm using a threshold of 2 nt [@Kozich2013; @Schloss2009]. ASVs were assigned to OTUs using a 3% distance threshold using mothur's cluster function with the OptiClust algorithm [@Westcott2017]. To minimize the effects of uneven sampling effort, samples were rarefied to the number of sequences in the smallest sample for each dataset. Because metrics of alpha diversity did not consistently follow a normal distribution, I used the non-parametric Wilcoxon rank test as implemented in R. For comparisons of Bray-Curtis distances, the amova function within mothur was used, which implements the analysis of molecular variance algorithm [@Anderson2001].

**Reproducibility.** All analyses were performed using mothur (version 1.44.1) and R (version `r paste(R.version$major, R.version$minor, sep='.')`) with the tidyverse (v.`r packageVersion("tidyverse")`), broom (v.`r packageVersion("broom")`), data.table (v.`r packageVersion("data.table")`), and cowplot (v.`r packageVersion("cowplot")`) packages. All of the code that was used in this analysis as well as the Makefile for running the analysis are available at https://github.com/SchlossLab/Schloss_Singletons_XXXXX_2019 along with the complete version history of the project.
